Regressions And Stuff

### 2016-09-30 paper2/paper.tex

\documentclass[10pt]{paper}
\usepackage{bm,amsmath,fullpage,graphicx,chngcntr}

\counterwithin{figure}{section}

\title{Regressions And Stuff}

@@

\maketitle

@@

\begin{figure}[ht!]
  \centering
  \label{fig:poly}
  \includegraphics[width=0.45\textwidth]{../images/poly0.png}
  \includegraphics[width=0.45\textwidth]{../images/poly1.png}
  \includegraphics[width=0.45\textwidth]{../images/poly3.png}
  \includegraphics[width=0.45\textwidth]{../images/poly10.png}
  \caption{Training data (green), generating curve (red), and predicted curve (blue) for various $n$.}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
  \label{fig:cos}
  \includegraphics[width=0.45\textwidth]{../images/cos2.png}
  \includegraphics[width=0.45\textwidth]{../images/cos8.png}
  \caption{Training data (green), generating curve (red), and predicted curve (blue) for various $n$.}
\end{figure}

@@

>>>>

\documentclass[10pt]{paper}
\usepackage{bm,amsmath,graphicx,chngcntr,subfigure}
\usepackage[margin=0.5in]{geometry}
\pagestyle{empty}
\counterwithin{figure}{section}
\pagenumbering{gobble}
\title{Regressions And Stuff}

@@

\maketitle
\section{Introduction}

In this paper, we discuss various methods of calculating regression. First, we discuss the benfits of using Stochastic Gradient Descent versus Batch Gradient Descent. Next, we analyze the performance of these methods in performing a Least Squares Linear regression. Finally, we discuss blah blah

\section{Gradient Descent}

\emph{Gradient descent} is an iterative optimization algorithm; in other words, it calculates the parameters $\mathbf{x}$ that minimize a given objective function $f(\mathbf{x})$. The algorithm repeatedly translates an initial \emph{guess} $\mathbf{x}_0$ in a direction proportional to the negative \emph{gradient} $\nabla f(x)$.

In every step of the itreration, we update our guess as following:
\begin{align*}
\mathbf{x}_{n+1} = \mathbf{x} - \lambda \nabla f(\mathbf{x}_{n})
\end{align*}
where $\lambda$ is the \emph{step size} of the iteration. The algorithm terminates upon the following convergence condition: 
\begin{align*}
|f(\mathbf{x}_{n+1}) - f(\mathbf{x}_{n})| < \delta
\end{align*}
where $\delta$ is the convergence \emph{threshold}. Upon convergence, the algorithm returns a final guess of $\mathbf{x}_{\text{opt}} = \mathbf{x}_{n+1}$.

In this section, we compare two methods of gradient descent: \emph{batch gradient descent}, which computes a gradient over all samples for each iteration, and \emph{stochastic gradient descent}, which instead uses pointwise gradients.

To analyze the perforamnce of the different algorithms, we use the \emph{Gaussian function}:
\begin{align*}
f(x) = - \dfrac{1}{\sqrt{(2\pi)^n |\Sigma|}} \exp \left[-\dfrac{1}{2} (x - u)^T \Sigma^{-1} (x-u)\right]
\end{align*}
In addition, we use the \emph{quadratic bowl function}:
\begin{align*}
f(x) = \dfrac{1}{2} x^T A x - x^T b 
\end{align*}
Finally, we use the \emph{least squares error} function:
\begin{align*}
J(\theta) = |X\theta - y|^2
\end{align*}

\subsection{Batch Gradient Descent}

In batch gradient descent, for every iteration we calculate the gradient across the entire sample set. In this section, we analyze the affects of step size ($\lambda$), threshold ($\delta$), and initial guess $\mathbf{x}_{0}$ on the performance and accuracy of the gradient descent algorithm.

First, we note that step size has a significant impact on the convergence rate of the algorithm. We run gradient descent on an arbitrary quadratic bowl function. As we increase the value of $\lambda$ from $0.0001$ to $0.001$, the magnitude of the gradient decreases more quickly, and the algorithm converges faster. Note that in this figure, all of the gradient descents were run with a fixed size of 2000 iterations to illustrate the effects on the gradient over time.

In addition, the step size and threshold control the accuracy of the gradient descent. As step size decreases, we expect a less accurate result because the gradient descent will likely converge at the boundary of the threshold, which a larger step size might lead to convergence well past the threshold. In addition, as the threshold decreases, we expect a more accurate result because the algorithm needs to run for longer before reaching a stable equilibrium. To verify these claims, we run gradient descent using a two-dimensional Gaussian with mean $(10, 10)$. As we vary $\lambda$ and $\delta$, we plot the least-squares error between the returned value and the actual optimum of $(10, 10)$ (see Figure 2). Indeed, the error decreases as $\lambda$ increases and $\delta$ decreases.

Finally, we observe the affect of the initial case on the rate of convergence. When running gradient descent on a two-dimensional Gaussian with mean $(10, 10)$, one would expect that as the intial guess appraoches the optimal value, the convergence rate increases. Indeed, running the algorithm with $\lambda = 1000$ and $\delta = 10^{-11}$, we notice that the rate of convergence is much higher as the initial guess approaches the actual value (see Figure 3).

\begin{figure}[ht!]
  \centering
  \begin{subfigure}
    \centering
      \includegraphics[width=0.3\textwidth]{../images/quadratic_steps_vs_mag}
    % \caption{As the step size increases, the algorithm converges quicker.}
  \end{subfigure}
  \begin{subfigure}
    \centering
      \includegraphics[width=0.3\textwidth]{../images/lsq_error_gradient}
    % \caption{As $\lambda$ increases and $\delta$ decreases, error decreases.}
  \end{subfigure}
  \begin{subfigure}
    \centering
      \includegraphics[width=0.3\textwidth]{../images/guess_vs_mags_2}
    % \caption{As $\mathbf{x}_0$ approaches $(10, 10)$, the rate of convergence increases.}
  \end{subfigure}
\end{figure}

\subsection{Stochastic Gradient Descent}

Sometimes, it might be inefficient to compute a gradient over the entire dataset for every iteration, especially when the dataset is large. One way to address this is to use stochastic gradient descent, which uses a pointwise gradient instead. In this experiment, we use least-squares regression, which has a pointwise gradient of

\begin{align*}
J(\theta; x^{(i)}, y^{(i)}) = (x^{(i)}\theta_t - y^{(i)})^2
\end{align*}
On each iteration, we update our value as follows:
\begin{align*}
\theta_{t+1} = \theta_{t} - \nu_t \nabla_\theta J(\theta_t; x^{(i)}, y^{(i)})
\end{align*}
where $\nabla = (\tau_0 + t)^{-\kappa}$ is the \emph{learning rate}, and $\tau_0$ and $\kappa$ are constants.

The benefit of stochastic gradient descent is that computations are quicker, and thus the algorithm converges faster, with respect to both time and the number of iterations. However, this comes with a decrease in performance, because each iteration uses less information. Refer to Figure 4 for the difference in the magnitude in the gradient between the two methods: note that stochastic gradient descent is much noisier, and in the end, does not converge to as accurate of a value as batch gradient descent does.

\begin{figure}[ht!]
  \center
  \begin{subfigure}
    \centering
      \includegraphics[width=0.45\textwidth]{../images/stochastic_gradient_steps}
    % \caption{Compare to batch gradient descent, the magnitude of the gradient during stochastic gradient descent shows significant noise.}
  \end{subfigure}
  \begin{subfigure}
    \centering
      \includegraphics[width=0.45\textwidth]{../images/batch_gradient_steps}
    % \caption{The magnitude of the gradient during batch gradient descent is much smoother.}
  \end{subfigure}
\end{figure}

\subsection{Effect of Finite Gradient}

When the objective function doesn't have a closed form solution, one can approximate the gradient using a central finite difference. In other words, we have 
\begin{align*}
\nabla f(x) \approx \nabla_h[f](x) = \dfrac{f(x+h/2) - f(x-h/2)}{h}
\end{align*}
Therefore, we only need the objective function to approximate the gradient. Note that as $h$ approaches zero, this approaches the actual value of the gradient; and has $h$ approaches infinity this value diverges. Figures 4 and 5 illustrate the impact of $h$ on the accuracy of the gradient: even for relatively large $h$ (especially for the quadratic bowl), the approximation is accurate to several orders of magnitude.

\begin{figure}[ht!]
  \centering
  \begin{subfigure}
    \centering
      \begin{tabular}{c | c}
      $h$  & $\nabla_h[f](x)$ \\  \hline
    0.01 & $-3.1704 \cdot 10^{-7}$ \\
    0.1 & $-3.1704 \cdot 10^{-7}$ \\
    1.0 & $-3.1696 \cdot 10^{-7}$ \\
    10.0 & $-3.0923 \cdot 10^{-7}$ \\
    100.0 & $-2.6198 \cdot 10^{-8}$ \\
    1000.0 & $\approx 0$ \\
      \end{tabular}
    % \caption{As $h$ increases, the accuracy of the central finite difference for this Gaussian decreases. The actual gradient value is $-3.1704 \cdot 10^{-7}$.}
  \end{subfigure}
  \qquad
  \begin{subfigure}
    \centering
      \begin{tabular}{c | c}
      $h$  & $\nabla_h[f](x)$ \\  \hline
    $10^{9}$ & $-280.000000$ \\
    $10^{11}$ & $-279.999991$ \\
    $10^{13}$ & $-280.000944$ \\
    $10^{15}$ & $-280.349077$ \\
    $10^{17}$ & $-299.759591$ \\
    $10^{19}$ & $\approx 0$ \\
      \end{tabular}
    % \caption{As $h$ increases, the accuracy of the central finite difference for this quadratic bowl decreases very slowly. The actual gradient value is $-280$.}
  \end{subfigure}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
  \label{fig:poly}
  \includegraphics[width=0.24\textwidth]{../images/poly0.png}
  \includegraphics[width=0.24\textwidth]{../images/poly1.png}
  \includegraphics[width=0.24\textwidth]{../images/poly3.png}
  \includegraphics[width=0.24\textwidth]{../images/poly10.png}
  \caption{Training data (green), generating curve (red), and predicted curve (blue) for $n = 0, 1, 3, 10$.}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
  \label{fig:cos}
  \includegraphics[width=0.3\textwidth]{../images/cos2.png}
  \includegraphics[width=0.3\textwidth]{../images/cos5.png}
  \includegraphics[width=0.3\textwidth]{../images/cos8.png}
  \caption{Training data (green), generating curve (red), and predicted curve (blue) for $n=2, 5, 8$.}
\end{figure}

@@

\subsection{Continuation of Simple Example}

In this subsection we apply ridge regression to the same example as the previous section, in which linear regression showed some overfitting. Varying $\lambda$ and $n$ gives us the plots in Figure \ref{fig:ridge}.

\begin{figure}[ht!]
  \centering
  \label{fig:ridge}
  \includegraphics[width=0.24\textwidth]{../images/ridge3low.png}
  \includegraphics[width=0.24\textwidth]{../images/ridge3high.png}
  \includegraphics[width=0.24\textwidth]{../images/ridge10low.png}
  \includegraphics[width=0.24\textwidth]{../images/ridge10high.png}
  \caption{Training data (green), generating curve (red), and predicted curve (blue) for $\lambda=0.0001, 0.1$ and $n = 3, 10$.}
\end{figure}

We can see that a good value of $\lambda$ gives a decent fit, while $\lambda$ too large results in \emph{underfitting}, where the prediction is not sensitive enough to the training data.

\subsection{Training and Validation}

As we just illustrated, choosing a good value of $\lambda$ is crucial for generalizing well. We now present a process for finding a good value for $\lambda$ (as well as $n$). Such a process is called \emph{model selection}.

We divide the given labeled data into two groups, one called the training set and one called the validation set. We vary $n$ and $\lambda$, and for each combination we train as above on the training set and compute SSE on the validation set. Finally, we select the combination of parameters that resulted in the smallest validation SSE.

To illustrate in an example, we have three sets of data: set A, set B, and a validation set. We perform the above process twice, once training on A and testing on B, and once vice versa. In both cases we try a variety of values of $\lambda$ and $n$ then evaluate the SSE on the validation set. The results are given in Figure \ref{fig:validationtable}
\begin{figure}[ht!]
  \centering
  \label{fig:validationtable}
  \begin{tabular}{c|cccccc}
    & 1e-05 & 0.0001 & 0.001 & 0.01 & 0.1 \\ \hline
  0 & 62.00 & 62.00 & 62.00 & 62.01 & 62.06 \\
  1 & 2.90 & 2.90 & 2.90 & 2.90 & 2.94 \\
  2 & 2.35 & \textbf{2.35} & 2.35 & 2.36 & 2.43 \\
  3 & 2.37 & 2.37 & 2.37 & 2.39 & 2.55 \\
  4 & 3.58 & 3.58 & 3.58 & 3.59 & 3.67 \\
  5 & 4.14 & 4.14 & 4.14 & 4.12 & 3.97 \\
  6 & 3.98 & 3.98 & 3.98 & 3.96 & 3.89 \\
  7 & 4.07 & 4.07 & 4.06 & 4.03 & 3.89 \\
  \end{tabular}
  \qquad
  \begin{tabular}{c|ccccc}
    & 0 & 0.3 & 0.6 & 0.9 & 1.2 \\ \hline
  0 & 72.44 & 70.95 & 69.63 & 68.47 & 67.44 \\
  1 & 35.19 & 34.09 & 33.14 & 32.34 & 31.67 \\
  2 & 38.53 & 38.65 & 38.76 & 38.87 & 38.96 \\
  3 & 30.30 & 28.46 & \textbf{28.09} & 28.47 & 29.21 \\
  4 & 130.20 & 67.54 & 51.63 & 46.17 & 44.35 \\
  5 & 625.43 & 368.40 & 287.26 & 242.74 & 212.14 \\
  6 & 680.05 & 533.60 & 481.26 & 440.20 & 406.68 \\
  7 & 492.85 & 360.39 & 448.42 & 544.91 & 634.85 \\
  \end{tabular}
  \caption{SSE values on validation set after training with various $n$ (rows) and $\lambda$ (columns). Left: training on set A. Right: training on set B. Best values bolded.}
\end{figure}
The SSE for the first case with $n = 2$ and $\lambda = 0.0001$, tested on set B, is $25.752$. In the second case, tested on set A with $n = 3$ and $\lambda = 0.6$, the SSE is $40.087483697$. Testing with parameters $n=4$ and $\lambda = 0.1$ for both cases, which the table would predict to be suboptimal, indeed gives bigger test SSEs of 31.1388659319 and 64.9392925156, respectively.

We can see in this example that the choice of training set matters. Also, we note that the best values for the parameters $n$ and $\lambda$ depends heavily on the given data.
